Estimating Animal Abundance with N-Mixture Models Using the \pkg{R-INLA} Package for R

### 2017-04-21 meehan_et_al_inla_nmix_jss/article.tex
Ready to send off after you look it over
@hrue , @nicolelmichel
Hi Haavard, Nicole.  I have fine tuned the R-INLA nMix manuscript.  In
tuning, I hope I haven't messed anything up.  If you have time, please
look it over one more time.  After I get your approval, I will send it
off for review.  The manuscript is now in LaTeX format.  The latest
version is also compiled as a PDF for convenience. Feel free to work
with either one. The LaTeX format follow the JSS style.  Seems like this
is an appropriate outlet for this manuscript.  Let me know if you have
other preferences. Also, let me know if yo

%% an abstract and keywords
\Abstract{
  Successful management of wildlife populations requires accurate estimates of abundance.  Abundance estimates can be confounded by imperfect detection during wildlife surveys. Development of N-mixture models has facilitated quantification of detectability and has improved our ability to produce accurate abundance estimates. The purpose of this study was to demonstrate the use of the \pkg{R-INLA} package to analyze N-mixture models and to compare performance of \pkg{R-INLA} to two other common approaches -- \proglang{JAGS} (via the \pkg{runjags} package), which uses Markov chain Monte Carlo methods and allows Bayesian inference, and \pkg{unmarked}, which uses Maximum Likelihood and allows Frequentist inference. Our results suggest that \pkg{R-INLA} is an attractive option for analyzing N-mixture models when (1) survey level covariates are not essential, (2) familiar model syntax and data format are desired (relative to other R packages), (3) fast computing time is necessary (\pkg{R-INLA} is 10 times faster than \pkg{unmarked}, 300 times faster than \proglang{JAGS}), and (4) Bayesian inference is preferred.

}
\Keywords{wildlife, abundance, detection, N-mixture models, \pkg{R-INLA}, \pkg{unmarked}, \proglang{R}, \proglang{JAGS}}
%% without formatting
\Plainkeywords{wildlife, abundance, detection, N-mixture models, R-INLA, R, BUGS}

@@

%% introduction
\section[Introduction]{Introduction}
\subsection[Background]{Background}
Successful management of wildlife species requires accurate estimates of abundance \citep{Yoccoz_Nichols_Boulinier_2001}.  One common method for estimating animal abundance is direct counts \citep{Pollock_Nichols_Simons_Farnsworth_Bailey_Sauer_2002}.  Efforts to obtain accurate abundance estimates via direct counts can be hindered by both the cryptic nature of many wildlife species and by other factors such as observer expertise, weather, and habitat structure.  The lack of perfect detection in wildlife surveys is common, and can cause abundance to be greatly underestimated \citep{Wenger_Freeman_2008, Joseph_Elkin_Martin_Possingham_2009}.

In recent years, new sampling schemes and modeling approaches have enabled improved estimates of animal abundance that are less biased by non-detection \citep{Denes_Silveira_Beissinger_2015}.  One such sampling scheme, termed a "metapopulation design" \citep{Kery_Royle_2010}, involves repeat visits in rapid succession to each of multiple study sites in a study area.  If, during repeat visits, the population is assumed to be closed (no immigration , emigration, or reproduction), then the ratio of detections to non-detections during repeated counts can inform an estimate of detection probability.  This detection probability can, in effect, be used to correct abundance estimates for imperfect detection \citep{Denes_Silveira_Beissinger_2015}.

@@

This estimation approach can be extended to cover $K$ distinct breeding seasons, which correspond with distinct years for annually-breeding wildlife species \citep{Kery_Dorazio_Soldaat_Van_Strien_Zuiderwijk_Royle_2009}.  In this case, population closure is assumed across $J$ surveys within year $k$, but is relaxed across years \citep{Kery_Dorazio_Soldaat_Van_Strien_Zuiderwijk_Royle_2009}. A simple specification of a multiple-year model is $N_{i,k} \sim \text{Poisson}(\lambda_{i,k}), \quad Y_{i,j,k} | N_{i,k} \sim \text{binomial}(N_{i,k}, p_{i,k})$.  Like the single-year specification, $\lambda$ is commonly modeled using site and site-year covariates, and $p$ using site-survey-year covariates.  There are other variations of N-mixture models that accommodate overdispersed counts through use of a negative binomial distribution \citep{Kery_Royle_2010}, zero-inflated Poisson distribution \citep{Wenger_Freeman_2008}, or survey-level random intercepts \citep{Kery_Schaub_2011}.  Yet other variations account for non-independent detection probabilities through use of a beta-binomial distribution \citep{Martin_Royle_Mackenzie_Edwards_Kery_Gardner_2011}, parse different components of detection through the use of unique covariates \citep{O'Donnell_Thompson_III_Semlitsch_2015}, or relax assumptions of population closure \citep{Chandler_Royle_King_2011, Dail_Madsen_2011}.  We do not discuss all of these variations here, but refer interested readers to \cite{Denes_Silveira_Beissinger_2015} for a nice overview.

The development of metapopulation designs and N-mixture models represents a significant advance in quantitative wildlife ecology.  However, there are practical issues that sometimes act as barriers to adoption.  Much of the development of N-mixture models, and many of the examples in the wildlife literature, implement models using Bayesian modeling software such as \proglang{WinBUGS}, \proglang{OpenBUGS}, or \proglang{JAGS} \citep{Lunn_Jackson_Best_Thomas_Spiegelhalter_2012}.  These are extremely powerful and flexible platforms for analyzing Bayesian models, but they come with a few important challenges.  First, many wildlife biologists are not accustomed to coding statistical models using the \proglang{BUGS} modeling syntax.  While there are several outstanding resources aimed at teaching this skill \citep{Royle_Dorazio_2008, Kery_2010, Kery_Schaub_2011, Kery_Royle_2015} it is, nonetheless, a considerable commitment.  Second, while Markov chain Monte Carlo (MCMC) chains converge quickly for relatively simple N-mixture models, convergence for more complex models can take hours to days, and may not occur at all.

There are other tools available for analyzing N-mixture models that alleviate some of these practical issues.  The \pkg{unmarked} package \citep{Fiske_Chandler_others_2011} for \proglang{R} statistical computing software \citep{R_Core_Team_2016} offers several options for analyzing N-mixture models within a Frequentist framework, with the capacity to accommodate overdispersion and dynamic populations.  The model coding syntax used in \pkg{unmarked} is a simple extension of the standard \proglang{R} syntax.  Models are analyzed using a Maximum Likelihood (ML), so model analysis is often completed in a fraction of the time taken using an MCMC approach.  The familiar model syntax and rapid model evaluation of \pkg{unmarked} has undoubtedly contributed to the broader adoption of N-mixture models by wildlife biologists.  However, it comes at a cost -- the loss of the intuitive inferential framework associated with Bayesian analysis.

Here we discuss analysis of N-mixture models using the \pkg{R-INLA} package \citep{Rue_Martino_Lindgren_Simpson_Riebler_2013} for \proglang{R}.  The \pkg{R-INLA} package uses integrated nested Laplace approximation (INLA) to derive posterior distributions for a large class of Bayesian statistical models that can be formulated as latent Gaussian models \citep{Rue_Martino_Chopin_2009}. INLA was developed to allow estimation of posterior distributions in a fraction of the time taken by MCMC.  Like \pkg{unmarked}, the model syntax used in the \pkg{R-INLA} package is a straightforward extension of the modeling syntax commonly used in R.  Also, like \pkg{unmarked}, the computational cost of analyzing models with \pkg{R-INLA} is relatively low.  The \pkg{R-INLA} approach is different from \pkg{unmarked} in that inference about model parameters falls within a Bayesian framework.

\subsection[Overall objectives]{Overall objectives}
The purpose of this manuscript is to demonstrate analysis of N-mixture models using the \pkg{R-INLA} package. In the process, we employ simulated and real count datasets, and analyze them using \proglang{JAGS}, via the \pkg{runjags} package \citep{Denwood_2016} for \proglang{R}, the \pkg{unmarked} package, and the \pkg{R-INLA} package. In each case, we demonstrate how data are formatted, how models are specified, how model estimates compare to simulation inputs, and how methods compare in terms of computational performance. We also explore a limitation of the \pkg{R-INLA} approach related to model specification.  Specifically, while it is possible to specify survey level covariates for detection using \proglang{JAGS} and \pkg{unmarked}, this is not possible using \pkg{R-INLA}.  Rather, survey level covariates of detection must be averaged to the site or site-year level.  Using an averaged detection covariate does allow accounting for site-level differences in survey conditions should they occur.  However, in the process of averaging, important information related to detection is discarded, which could lead to biased detection and abundance estimates under certain conditions.

@@

\subsection[Simulated data]{Simulated data}
The data simulated for this analysis were intended to represent a typical wildlife study. To put the simulation in context, consider an effort to estimate the abundance of a bird species in a national park, within which are located 72 study sites. At each site, 3 replicate surveys are conducted within 6 weeks, during the peak of the breeding season when birds are likely to be singing. In order to estimate a trend in abundance over time, these repeated surveys are conducted over a 9-year period.

@@

As is commonly the case, we assume that the counts are overdispersed due to the effects of unknown variables. Overdispersion was generated and modeled using a negative binomial distribution for the count component of the model. The specific model parameters and their simulated values were: p.b0 = 1.0 ($p$ intercept), p.b1 = -2.0 (effect of $p$ covariate 1), p.b3 = 1.0 (effect of $p$ covariate 3), lam.b0 = 2.0 ($\lambda$ intercept), lam.b1 = 2.0 (effect of $\lambda$ covariate 1), lam.b2 = -3.0 (effect of $\lambda$ covariate 2), lam.b4 = 1.0 (effect of year on $\lambda$), disp.size = 3.0 (size of the overdispersion parameter). All independent variables in the simulation were centered at zero to alleviate computational difficulties and to make model intercepts more easily interpreted.

@@

%% example 1
\section[Example I]{Example I}
\subsection[Goals]{Goals}
In Example I, we demonstrate the use of \pkg{R-INLA} and compare use and performance to similar analyses using \proglang{JAGS} and \pkg{unmarked}.  In this exercise, the functional forms of \proglang{JAGS}, \pkg{unmarked}, and \pkg{R-INLA} models match the data generating process.  Specifically, we used the covariate x.p.3.mean to generate the count matrix y1 and analyzed the data with models that use x.p.3.mean as a covariate.  This exercise was intended to demonstrate the differences and similarities in use, computation time, and estimation results across the three packages when the specified model was reasonably close to the data generating process.

@@

\begin{Code}
R> jags.model.string <- "
+    model {
+    # priors
+    intP ~ dnorm(0, 0.01)       # detection intercept
+    bCov1P ~ dnorm(0, 0.01)     # detection cov 1 effect
+    bCov3P ~ dnorm(0, 0.01)     # detection cov 3 effect
+    intLam ~ dnorm(0, 0.01)     # lambda intercept
+    bCov1Lam ~ dnorm(0, 0.01)   # lambda cov 1 effect
+    bCov2Lam ~ dnorm(0, 0.01)   # lambda cov 2 effect
+    bYr ~ dnorm(0, 0.01)        # year effect
+    overDisEst ~ dunif(0, 5)    # overdispersion size
+    # abundance component
+    for (k in 1:nYears){
+      for (i in 1:nSites){
+        N[i, k] ~ dnegbin(prob[i, k], overDisEst) # negative binomial specification
+        prob[i, k] <- overDisEst / (overDisEst + lambda[i, k]) # overdispersion effect
+        log(lambda[i, k]) <- intLam + (bCov1Lam * x.lam.1[i]) + (bCov2Lam * x.lam.2[i]) +
+                          (bYr * yr[k])
+    # detection component
+        for (j in 1:nSurveys){
+          y[i, j, k] ~ dbin(p[i,j,k], N[i,k])
+          p[i, j, k] <- exp(lp[i,j,k]) / (1 + exp(lp[i,j,k]))
+          lp[i, j, k] <- intP + (bCov1P * x.p.1[i]) + (bCov3P * x.p.3[i, k])
+        } # close j loop
+      } # close i loop
+    } # close k loop
+  } # close model loop
+  "
\end{Code}

@@

R> params <- c("intP", "bCov1P", "bCov3P", "intLam", "bCov1Lam","bCov2Lam",
+              "bYr", "overDisEst") # parameters to monitor
R> jags.data <- list(y = sim.data$y1, x.lam.1 = sim.data$x.lam.1,
+               x.lam.2 = sim.data$x.lam.2, yr = sim.data$yr,
+               x.p.1 = sim.data$x.p.1, x.p.3 = sim.data$x.p.3.mean,
+               nSites = sim.data$n.sites, nSurveys = sim.data$n.surveys,
+               nYears = sim.data$n.years)
R> N.init <- sim.data$y1 # initial count values
R> N.init[is.na(N.init)] <- 1 # clean up NA's
R> N.init <- apply(N.init, c(1, 3), max) + 1 # zero values cause trouble
R> inits <- function() list(N = N.init, intLam = rnorm(1, 0, 0.01),
+                      intP = rnorm(1, 0, 0.01), bCov1P = rnorm(1, 0, 0.01),
+                      bCov2Lam = rnorm(1,0,0.01), bCov1Lam = rnorm(1, 0, 0.01),
+                      bCov3P = rnorm(1, 0, 0.01), bYr = rnorm(1, 0, 0.01),
+                      overDisEst = runif(1, 0.5, 2.5))
\end{Code}

Finally, we set the run parameters and started the MCMC process. Run parameters were chosen such that MCMC diagnostics indicated converged chains (Gelman-Rubin statistics < 1.05) and reasonably robust posterior distributions (effective sample sizes > 1000).  Note that the recommended number of effective samples for particularly robust inference is closer to 6000 \citep{Gong_Flegal_2016}.  So MCMC processing times reported here could be considered low estimates.

\begin{Code}
R> nc <- 3; na <- 2500; nb <- 2500; ni <- 5000; nt <- 10 # set run parameters
R> out.jags <- run.jags(model = jags.model.string, data = jags.data,
+                       monitor = params, n.chains = nc, inits = inits,
+                       burnin = nb, adapt = na, sample = ni, thin = nt,
+                       modules = "glm on", method = "parallel")
\end{Code}

Mean parameter estimates from the \proglang{JAGS} model were reasonably close to, and not significantly different from, the input values used to generate the data (Fig. \ref{fig:fig1}).  The potential scale reduction factor for all variables was < 1.01, and the effective sample size for all variables was > 2410.  The simulation ran in parallel on 3 virtual cores, 1 MCMC chain per core, and took approximately 2104 seconds.

@@

Maximum likelihood estimates from the \pkg{unmarked} analysis are also close to, and not significantly different from, input values (Fig. \ref{fig:fig1})  The \pkg{unmarked} estimates were produced in approximately 77 seconds.

@@

The call to the inla() function includes several components.  First is the model statement.  On the left side of the formula is the counts.and.count.covs object that includes the matrix of counts and the covariates related to $\lambda$.  On the right side of the formula is a 1 to specify a global intercept for $p$ and the two fixed covariates for $p$.  Note that a wide range of random effects for $p$ could be added to the right side of the formula \citep{Rue_Martino_Lindgren_Simpson_Riebler_2013}.  The second argument describes the data, provided here as a list that corresponds with the model formula.  Third is the likelihood family, which can take values of "nmix" for a Poisson-binomial mixture and "nmixnb" for a negative binomial-binomial mixture.  The fourth and fifth arguments specify the priors for the two model components.  In this case, the priors are specified similarly to those in the \proglang{JAGS} model, but a variety of other options are available. Several other characteristics of the analysis can be modified using the call to inla().  See \cite{Rue_Martino_Lindgren_Simpson_Riebler_2013} for details.

\begin{Code}
R> out.inla <- inla(counts.and.count.covs ~ 1 + x.p.1.inla + x.p.3.inla,
+           data = list(counts.and.count.covs = counts.and.count.covs,
+                       x.p.1.inla = x.p.1.inla, x.p.3.inla = x.p.3.inla),
+           family = "nmixnb",
+           control.fixed = list(mean = 0, mean.intercept = 0, prec = 0.01,
+                                prec.intercept = 0.01),
+           control.family = list(hyper = list(theta1 = list(param = c(0, 0.01)),
+                                              theta2 = list(param = c(0, 0.01)),
+                                              theta3 = list(param = c(0, 0.01)),
+                                              theta4 = list(param = c(0, 0.01)),
+                                              theta5 = list(prior = "flat",
+                                                            param = numeric()))))
\end{Code}

Mean parameter estimates from the \pkg{R-INLA} model are also close to, and not significantly different from, input values (Fig. \ref{fig:fig1})  Full marginal posterior distributions from \pkg{R-INLA} are shown in Fig. \ref{fig:fig1}.  The analysis took approximately 6 seconds.

\subsection[Example I summary]{Example I summary}
Example I demonstrated basic use of \pkg{R-INLA} and highlighted similarities and differences between it and two other commonly used approaches.  In demonstrating the use of \pkg{R-INLA}, we show that the input data format is not complicated, and that the formatting process can be accomplished with relatively few lines of code.  Similarly, model specification uses a straightforward extension of the standard syntax in R, where the count matrix and covariates for $\lambda$ are specified through an \pkg{R-INLA} object included on the left side of the formula, and covariates for $p$ are specified on the right side of the formula. Data format and model specification syntax of \pkg{R-INLA} are not too different from \pkg{unmarked}, which are both considerably different from \proglang{JAGS} and other \proglang{BUGS}-oriented MCMC software.

Regarding performance, \pkg{R-INLA}, \pkg{unmarked}, and \proglang{JAGS} successfully extracted simulation input values.  Fig. \ref{fig:fig1} depicts marginal posterior distributions produced by \proglang{JAGS} and \pkg{R-INLA}, and 95\% confidence intervals from \pkg{unmarked}.   These results derive from data resulting from one random manifestation of the input values.  Thus, we do not expect the posterior distributions for the estimates to be centered at the input values, which would be expected if the simulation was repeated many times.  However, we do expect the input values to fall somewhere within the posterior distributions and 95\% confidence intervals, which is what occurs here.  Fig. \ref{fig:fig1} shows that, for similarly specified models, \pkg{R-INLA} (dashed black lines) and \proglang{JAGS} (solid gray lines) yielded practically identical marginal posterior distributions for model parameters.

>>>>

%% an abstract and keywords
\Abstract{
  Successful management of wildlife populations requires accurate estimates of abundance. Abundance estimates can be confounded by imperfect detection during wildlife surveys. N-mixture models enable quantification of detection probability and often produce abundance estimates that are less biased. The purpose of this study was to demonstrate the use of the \pkg{R-INLA} package to analyze N-mixture models and to compare performance of \pkg{R-INLA} to two other common approaches -- \proglang{JAGS} (via the \pkg{runjags} package), which uses Markov chain Monte Carlo and allows Bayesian inference, and \pkg{unmarked}, which uses Maximum Likelihood and allows Frequentist inference. We show that \pkg{R-INLA} is an attractive option for analyzing N-mixture models when (1) familiar model syntax and data format (relative to other \proglang{R} packages) are desired, (2) survey level covariates of detection are not essential, (3) fast computing times are necessary (\pkg{R-INLA} is 10 times faster than \pkg{unmarked}, 300 times faster than \proglang{JAGS}), and (4) Bayesian inference is preferred.
}
\Keywords{abundance, detection, \proglang{JAGS}, N-mixture model, \proglang{R}, \pkg{R-INLA}, \pkg{unmarked}, wildlife}
%% without formatting
\Plainkeywords{wildlife, abundance, detection, N-mixture models, R-INLA, R, BUGS}

@@

%% introduction
\section[Introduction]{Introduction}
\subsection[Background]{Background}
Successful management of wildlife species requires accurate estimates of abundance \citep{Yoccoz_Nichols_Boulinier_2001}. One common method for estimating animal abundance is direct counts \citep{Pollock_Nichols_Simons_Farnsworth_Bailey_Sauer_2002}. Efforts to obtain accurate abundance estimates via direct counts can be hindered by the cryptic nature of many wildlife species and by other factors such as observer expertise, weather, and habitat structure \citep{Denes_Silveira_Beissinger_2015}. The lack of perfect detection in wildlife surveys is common, and can cause abundance to be greatly underestimated \citep{Wenger_Freeman_2008, Joseph_Elkin_Martin_Possingham_2009}.

In recent years, new sampling schemes and modeling approaches have enabled improved estimates of animal abundance that are less biased by non-detection \citep{Denes_Silveira_Beissinger_2015}. One such sampling scheme, termed a metapopulation design \citep{Kery_Royle_2010}, involves repeat visits in rapid succession to each of multiple study sites in a study area. If, during repeat visits, the population is assumed to be closed (no immigration , emigration, or reproduction; static abundance), then the ratio of detections to non-detections during repeated counts can inform an estimate of detection probability. This detection probability can, in effect, be used to correct abundance estimates for imperfect detection \citep{Royle_2004}.

@@

This estimation approach can be extended to cover $K$ distinct breeding seasons, which correspond with distinct years for wildlife species that breed annually \citep{Kery_Dorazio_Soldaat_Van_Strien_Zuiderwijk_Royle_2009}. In this case, population closure is assumed across $J$ surveys within year $k$, but is relaxed across years \citep{Kery_Dorazio_Soldaat_Van_Strien_Zuiderwijk_Royle_2009}. A simple specification of a multiple-year model is $N_{i,k} \sim \text{Poisson}(\lambda_{i,k}), \quad Y_{i,j,k} | N_{i,k} \sim \text{binomial}(N_{i,k}, p_{i,k})$. Like the single-year specification, $\lambda$ is commonly modeled using site and site-year covariates, and $p$ using site-survey-year covariates. There are other variations of N-mixture models that accommodate overdispersed counts through use of a negative binomial distribution \citep{Kery_Royle_2010}, a zero-inflated Poisson distribution \citep{Wenger_Freeman_2008}, or survey-level random intercepts \citep{Kery_Schaub_2011}. Yet other variations account for non-independent detection probabilities through use of a beta-binomial distribution \citep{Martin_Royle_Mackenzie_Edwards_Kery_Gardner_2011}, parse different components of detection through the use of unique covariates \citep{O'Donnell_Thompson_III_Semlitsch_2015}, or relax assumptions of population closure \citep{Chandler_Royle_King_2011, Dail_Madsen_2011}. We do not discuss all of these variations here, but refer interested readers to \cite{Denes_Silveira_Beissinger_2015} for an overview.

The development of metapopulation designs and N-mixture models represents a significant advance in quantitative wildlife ecology. However, there are practical issues that sometimes act as barriers to adoption. Many of the examples of N-mixture models in the wildlife literature have employed Bayesian modeling software such as \proglang{WinBUGS}, \proglang{OpenBUGS}, or \proglang{JAGS} \citep{Lunn_Jackson_Best_Thomas_Spiegelhalter_2012}. These are extremely powerful and flexible platforms for analyzing hierarchical models, but they come with a few important challenges.  First, many wildlife biologists are not accustomed to coding statistical models using the \proglang{BUGS} modeling syntax. While there are several outstanding resources aimed at teaching this skill \citep{Royle_Dorazio_2008, Kery_2010, Kery_Schaub_2011, Kery_Royle_2015} it is, nonetheless, a considerable commitment. Second, while Markov chain Monte Carlo (MCMC) chains converge quickly for relatively simple N-mixture models, convergence for more complex models can take hours to days, and may not occur at all.

There are other tools available for analyzing N-mixture models that alleviate some of these practical issues.  The \pkg{unmarked} package \citep{Fiske_Chandler_others_2011} for \proglang{R} statistical computing software \citep{R_Core_Team_2016} offers several options for analyzing N-mixture models within a Frequentist framework, with the capacity to accommodate overdispersion and dynamic populations. The model coding syntax used in \pkg{unmarked} is a simple extension of the standard \proglang{R} syntax. Models are analyzed using a Maximum Likelihood (ML), so model analysis is often completed in a fraction of the time taken using an MCMC approach. The familiar model syntax and rapid model evaluation of \pkg{unmarked} has undoubtedly contributed to the broader adoption of N-mixture models by wildlife biologists. However, it comes at a cost -- loss of the intuitive inferential framework associated with Bayesian analysis.

Here we discuss analysis of N-mixture models using the \pkg{R-INLA} package \citep{Rue_Martino_Lindgren_Simpson_Riebler_2013} for \proglang{R}. The \pkg{R-INLA} package uses integrated nested Laplace approximation (INLA) to derive posterior distributions for a large class of Bayesian statistical models that can be formulated as latent Gaussian models \citep{Rue_Martino_Chopin_2009}. INLA was developed to allow estimation of posterior distributions in a fraction of the time taken by MCMC. Like \pkg{unmarked}, the model syntax used in the \pkg{R-INLA} package is a straightforward extension of the modeling syntax commonly used in \proglang{R}. Also, like \pkg{unmarked}, the computational cost of analyzing models with \pkg{R-INLA} is relatively low. The \pkg{R-INLA} approach is different from \pkg{unmarked} in that inference about model parameters falls within a Bayesian framework.

\subsection[Overall objectives]{Overall objectives}
The purpose of this manuscript is to demonstrate analysis of N-mixture models using the \pkg{R-INLA} package. In the process, we employ simulated and real count datasets, and analyze them using \proglang{JAGS}, via the \pkg{runjags} package \citep{Denwood_2016} for \proglang{R}, the \pkg{unmarked} package, and the \pkg{R-INLA} package. In each case, we demonstrate how data are formatted, how models are specified, how model estimates compare to simulation inputs, and how methods compare in terms of computational performance. We also explore a limitation of the \pkg{R-INLA} approach related to model specification. Specifically, while it is possible to specify survey level covariates for detection using \proglang{JAGS} and \pkg{unmarked}, this is not possible using \pkg{R-INLA}. Rather, survey level covariates of detection must be averaged to the site or site-year level. Using an averaged detection covariate does allow accounting for site-level differences in survey conditions, should they occur. However, in the process of averaging, information related to detection within a site or site-year combination is discarded, which could lead to biased detection and abundance estimates under certain conditions.

@@

\subsection[Simulated data]{Simulated data}
The data simulated for this analysis were intended to represent a typical wildlife population study. To put the simulation in context, consider an effort to estimate the abundance of a bird species in a national park, within which are located 72 study sites. At each site, 3 replicate surveys are conducted within 6 weeks, during the peak of the breeding season, when birds are likely to be singing. In order to estimate a trend in abundance over time, clusters of repeated surveys are conducted each breeding season over a 9-year period.

@@

As is commonly the case, we assume that the counts are overdispersed due to the effects of unknown variables. Overdispersion was generated and modeled using a negative binomial distribution for the count component of the model. The specific model parameters and their simulated values were: p.b0 = 1.0 (logit($p$) intercept), p.b1 = -2.0 (effect of covariate 1 on logit($p$)), p.b3 = 1.0 (effect of covariate 3 on logit($p$)), lam.b0 = 2.0 (log(lambda) intercept), lam.b1 = 2.0 (effect of covariate 1 on log(lambda)), lam.b2 = -3.0 (effect of covariate 2 on log(lambda)), lam.b4 = 1.0 (effect of year on log(lambda)), disp.size = 3.0 (size of the overdispersion parameter). All independent variables in the simulation were centered at zero to alleviate computational difficulties and to make model intercepts more easily interpreted.

@@

%% example 1
\section[Example I]{Example I}
\subsection[Goals]{Goals}
In Example I, we demonstrate the use of \pkg{R-INLA} and compare use and performance to similar analyses using \proglang{JAGS} and \pkg{unmarked}. In this exercise, the functional forms of \proglang{JAGS}, \pkg{unmarked}, and \pkg{R-INLA} models match the data generating process.  Specifically, we used the covariate x.p.3.mean to generate the count matrix y1 and analyzed the data with models that use x.p.3.mean as a covariate. This exercise was intended to demonstrate the differences and similarities in use, computation time, and estimation results across the three methods when the specified model was reasonably close to the data generating process.

@@

\begin{Code}
R> jags.model.string <- "
+    model {
+    # priors
+    intP ~ dnorm(0, 0.01)       # detection intercept
+    bCov1P ~ dnorm(0, 0.01)     # detection cov 1 effect
+    bCov3P ~ dnorm(0, 0.01)     # detection cov 3 effect
+    intLam ~ dnorm(0, 0.01)     # lambda intercept
+    bCov1Lam ~ dnorm(0, 0.01)   # lambda cov 1 effect
+    bCov2Lam ~ dnorm(0, 0.01)   # lambda cov 2 effect
+    bYr ~ dnorm(0, 0.01)        # year effect
+    overDisEst ~ dunif(0, 5)    # overdispersion size
+    # abundance component
+    for (k in 1:nYears){
+      for (i in 1:nSites){
+        N[i, k] ~ dnegbin(prob[i, k], overDisEst) # negative binomial
+        prob[i, k] <- overDisEst / (overDisEst + lambda[i, k])
+        log(lambda[i, k]) <- intLam + (bCov1Lam * x.lam.1[i]) + 
+                             (bCov2Lam * x.lam.2[i]) + (bYr * yr[k])
+    # detection component
+        for (j in 1:nSurveys){
+          y[i, j, k] ~ dbin(p[i,j,k], N[i,k])
+          p[i, j, k] <- exp(lp[i,j,k]) / (1 + exp(lp[i,j,k]))
+          lp[i, j, k] <- intP + (bCov1P * x.p.1[i]) + (bCov3P * x.p.3[i, k])
+        } # close j loop
+      } # close i loop
+    } # close k loop
+  } # close model loop
+  "
\end{Code}

@@

\begin{Code}
R> params <- c("intP", "bCov1P", "bCov3P", "intLam", "bCov1Lam","bCov2Lam",
+              "bYr", "overDisEst") # parameters to monitor
R> jags.data <- list(y = sim.data$y1, x.lam.1 = sim.data$x.lam.1,
+               x.lam.2 = sim.data$x.lam.2, yr = sim.data$yr,
+               x.p.1 = sim.data$x.p.1, x.p.3 = sim.data$x.p.3.mean,
+               nSites = sim.data$n.sites, nSurveys = sim.data$n.surveys,
+               nYears = sim.data$n.years)
R> N.init <- sim.data$y1 # initial count values
R> N.init[is.na(N.init)] <- 1 # clean up NA's
R> N.init <- apply(N.init, c(1, 3), max) + 1 # zero values cause trouble
R> inits <- function() list(N = N.init, intLam = rnorm(1, 0, 0.01),
+               intP = rnorm(1, 0, 0.01), bCov1P = rnorm(1, 0, 0.01),
+               bCov2Lam = rnorm(1, 0, 0.01), bCov1Lam = rnorm(1, 0, 0.01),
+               bCov3P = rnorm(1, 0, 0.01), bYr = rnorm(1, 0, 0.01),
+               overDisEst = runif(1, 0.5, 2.5))
\end{Code}

Finally, we set the run parameters and started the MCMC process. Run parameters were chosen such that MCMC diagnostics indicated converged chains (potential scale reduction factors < 1.05) and reasonably robust posterior distributions (effective sample sizes > 1000). Note that the recommended number of effective samples for particularly robust inference is closer to 6000 \citep{Gong_Flegal_2016}. So MCMC processing times reported here could be considered low estimates.

\begin{Code}
R> nc <- 3; na <- 2500; nb <- 2500; ni <- 5000; nt <- 10 # run parameters
R> out.jags <- run.jags(model = jags.model.string, data = jags.data,
+                       monitor = params, n.chains = nc, inits = inits,
+                       burnin = nb, adapt = na, sample = ni, thin = nt,
+                       modules = "glm on", method = "parallel")
\end{Code}

Mean parameter estimates from the \proglang{JAGS} model were reasonably close to, and not significantly different from, the input values used to generate the data (Fig. \ref{fig:fig1}). The potential scale reduction factor for all variables was < 1.01, and the effective sample size for all variables was > 2409. The simulation ran in parallel on 3 virtual cores, 1 MCMC chain per core, and took approximately 2104 seconds.

@@

Maximum Likelihood estimates from the \pkg{unmarked} analysis are also close to, and not significantly different from, input values (Fig. \ref{fig:fig1})  The \pkg{unmarked} estimates were produced in approximately 77 seconds.

@@

The call to the inla() function includes several components. First is the model statement. On the left side of the formula is the counts.and.count.covs object that includes the matrix of counts and the covariates related to $\lambda$.  On the right side of the formula is a 1 to specify a global intercept for $p$ and the two covariates for $p$.  Note that a wide range of random effects for $p$ could be added to the right side of the formula using the f() syntax \citep{Rue_Martino_Lindgren_Simpson_Riebler_2013}. The second argument describes the data, provided here as a list that corresponds with the model formula. Third is the likelihood family, which can take values of "nmix" for a Poisson-binomial mixture and "nmixnb" for a negative binomial-binomial mixture. The fourth (control.fixed, for detection parameters) and fifth (control.family, for abundance and overdispersion parameters) arguments specify the priors for the two model components. In this case, the priors are specified similarly to those in the \proglang{JAGS} model, but a variety of other options are available. Several other characteristics of the analysis can be modified in the call to inla(). See \cite{Rue_Martino_Lindgren_Simpson_Riebler_2013} for details.

\begin{Code}
R> out.inla <- inla(counts.and.count.covs ~ 1 + x.p.1.inla + x.p.3.inla,
+           data = list(counts.and.count.covs = counts.and.count.covs,
+                       x.p.1.inla = x.p.1.inla, x.p.3.inla = x.p.3.inla),
+           family = "nmixnb",
+           control.fixed = list(mean = 0, mean.intercept = 0, prec = 0.01,
+                                prec.intercept = 0.01),
+           control.family = list(hyper = 
+                                  list(theta1 = list(param = c(0, 0.01)),
+                                       theta2 = list(param = c(0, 0.01)),
+                                       theta3 = list(param = c(0, 0.01)),
+                                       theta4 = list(param = c(0, 0.01)),
+                                       theta5 = list(prior = "flat",
+                                                     param = numeric()))))
\end{Code}

Mean parameter estimates from the \pkg{R-INLA} model are also close to, and not significantly different from, input values (Fig. \ref{fig:fig1}). Full marginal posterior distributions from \pkg{R-INLA} are shown in Fig. \ref{fig:fig1}. The analysis took approximately 6 seconds.

\subsection[Example I summary]{Example I summary}
Example I demonstrated basic use of \pkg{R-INLA} and highlighted similarities and differences between it and two other commonly used approaches. In demonstrating the use of \pkg{R-INLA}, we show that the input data format is not complicated, and that the formatting process can be accomplished with relatively few lines of code. Similarly, model specification uses a straightforward extension of the standard syntax in \proglang{R}, where the count matrix and covariates for $\lambda$ are specified through an \pkg{R-INLA} object included on the left side of the formula, and covariates for $p$ are specified on the right side of the formula. The data format and model specification syntax of \pkg{R-INLA} are not too different from \pkg{unmarked}, which are both considerably different from \proglang{JAGS} and other \proglang{BUGS}-oriented MCMC software.

Regarding performance, \pkg{R-INLA}, \pkg{unmarked}, and \proglang{JAGS} successfully extracted simulation input values. Fig. \ref{fig:fig1} shows marginal posterior distributions produced by \proglang{JAGS} and \pkg{R-INLA}, and estimates and 95\% confidence intervals from \pkg{unmarked}. These results derive from data from one random manifestation of the input values. Thus, we do not expect the posterior distributions for the estimates to be centered at the input values, which would be expected if the simulation was repeated many times. However, we do expect the input values to fall somewhere within the posterior distributions and 95\% confidence limits, which is what occurs here.  Fig. \ref{fig:fig1} shows that, for similarly specified models, \pkg{R-INLA} (dashed black lines) and \proglang{JAGS} (solid gray lines) yielded practically identical marginal posterior distributions for model parameters.

Where \pkg{R-INLA}, \pkg{unmarked}, and \proglang{JAGS} differed was in computing time.  In this example, \pkg{R-INLA} took 6 seconds, \pkg{unmarked} took 77 seconds, and \proglang{JAGS} took 2104 seconds to produce results. Thus, \pkg{R-INLA} was approximately 10 times faster than \pkg{unmarked} and 300 times faster than \proglang{JAGS}. This was the case despite the fact that \pkg{unmarked} produces ML estimates and the \proglang{JAGS} model was run in parallel with each of three MCMC chains simulated on a separate virtual computing core. If parallel computing had not been used with \proglang{JAGS}, processing the \proglang{JAGS} model would have taken approximately twice as long. If MCMC simulations were run until an effective sample size of 6000 was reached, processing times would have doubled again.
